plyranges: a grammar for transforming genomics data

### 2018-04-09 paper.Rmd
add in valr citation closes #8

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogenous genomics data heterogenous genomics data by expresssing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED.
An example of the former is the Genome Query Language (GQL) and its distributed 
implementation GenAp which use an SQL-like syntax for fast retrieval of 
information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. 
Similarly, the Genometric Query Language (GMQL) implements a 
relational algebra for combining genomic datasets @Kaitoua2017-pw.  
The command line application BEDtools develops an extensive algebra for 
performing arithmetic between two or more sets of genomic regions 
@Quinlan2010-gc; @Dale2011-js. All of the
aformentioned DSLs are designed to be evaluated either at the command
line or embedded in scripts for batch processing. They exist in a
sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@Lawrence2013-wg; @Huber2015-ei] define a DSL for analysing genomics data with 
R, an interactive data analysis environment that encourages
reproducibility and provides high-level abstractions for manipulating,
modeling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. The Bioconductor
infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behavior that often confuse the typical end
user.

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
@Wickham2014-jc. A tidy dataset is defined to be to a tabular
data structure that has observations as rows and columns as variables, and each
tidy dataset represents measurements from a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. The genomic interval data structure follows 
this pattern: it is a rectangular table corresponding to a single biological 
context. Each row contains a single observation and each column is a variable about
that observation.

>>>>

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogenous genomics data heterogenous genomics data by expresssing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED.An example of the former is the Genome Query 
Language (GQL) and its distributed implementation GenAp which use an 
SQL-like syntax for fast retrieval of information of unprocessed sequencing data @Kozanitis2014-va; @Kozanitis2016-bm. Similarly, the Genometric Query Language 
(GMQL) implements a relational algebra for combining genomic datasets 
@Kaitoua2017-pw.  The command line application BEDtools develops an extensive algebra for performing arithmetic between two or more sets of genomic regions 
@Quinlan2010-gc. All of the aformentioned DSLs are designed to be evaluated
either at the commandline or embedded in scripts for batch processing. They 
exist in asparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analysing genomics data with R, an interactive data analysis environment that 
encouragesreproducibility and provides high-level abstractions for manipulating,
modeling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include pybedtools and valr 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by in the aformentioned Bioconductor packages and are not easily 
extended.

The Bioconductor infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behavior that often confuse the typical end
user.

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
@Wickham2014-jc. A tidy dataset is defined to be to a tabular
data structure that has observations as rows and columns as variables, and each
tidy dataset represents measurements from a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. The `dplyr` pacakge [@Wickham2017-dplyr] 
defines an API that maps notions from the general relational algebra to 
operations on tidy data. It expresses each operation as a cohesive, endomorphic
verb. Taken together these features enable a user to write human
readable analysis workflows.
